Relationships in the extremes and their influence on competition and coexistence

### 2021-08-16 paper/Paper.Rnw
figures and ATA exlcusion/rescue definition and trait space stuff - preparing to send to Dan

To further explore the model and acheive G3 from the introduction i.e., defining the ATA rescue or ATA exlusion subset of trait space for which species will coexist or become competitively excluded in our model, we plotted the invasion growth rate, $r_{i/i}-q_{12}r_{i/j}$ and the invasion growth rate [IGR] without ATA contributribution  $r_{i/i}-q_{12}r_{i/j} - \Delta_i^{[EC]}$ against the difference between species, $\mu_1-\mu_2$. The same mathematical simplifications described above still hold, such that $\mu1$ and $\mu2$ are the only ways that the species differ and computations for coexistence in the lottery model only depend on the difference, $\mu_1-\mu_2$. We henceforward call this difference the trait difference between species. Letting the trait $\mu$ vary within a one dimensional space, this space is then a trait space. Letting there me two points $\mu_1$ and $\mu_2$ within that trait space, then the difference, $\mu_1-\mu_2$, is a subset of that trait space. We then define $\psi_{ATA}$ to be the subset of trait space that allows a postive IGR when ATA is included and $\psi_{~ATA}$ to be the subset of trait space that allows a postive IGR when ATA is not included. We then use symmetric difference $\psi_{ATA} \Delta \psi_{~ATA}$ to make the following conslusions. If the length/range? of $\psi_{ATA}$ > the length/range? of $\psi_{~ATA}$ then $\psi_{ATA} \Delta \psi_{~ATA}$ consitutes ATA rescue and the degree of species trait difference that allows ATA rescue to act, hence, the ATA rescue trait space. If the length of $\psi_{ATA}$ < the length/range? of $\psi_{~ATA}$ then $\psi_{ATA} \Delta \psi_{~ATA}$ consitutes ATA exclusion and the degree of species trait difference that allows ATA exclusion to act, hence, the ATA exclusion trait space. We highlight and color code these areas in our plots. We made the plots for values of $\delta=0.5, 0.8, 1.0$ and $\sigma=2.0,4.0,5.0, 6.0$. The spaces in between the lines where they intersect the x-axis at y=0, indicate either an ATA rescue trait space or an ATA exclusion trait space. If the line plotting IGR is to the left of the line plotting IGR without ATA, then this space is an ATA rescue trait space. If the line plotting IGR is to the right of the line plotting IGR without ATA, then this space is an ATA exclusion trait space.

@@

\begin{figure}
\caption{
%Figure displaying the noise. Three panels, left-, symmetric, and right-tail assoc noise. Use standard normal marginals.Explain in caption that alternatives for $(b_1,b_2)$ are obtained by multipling by $\sigma$ and then adding $(\mu_1,\mu_2)$.Jasmin to make and insert the figure here using the includegraphics command, and then draft a caption along the lines described here.
Simple example of bivariate noise with standard normal marginals. A) is left-tail associated, B) is symmetric, and C) is right-tail associated. All three cases have the same Pearson correlation of $\rho=$, up to sampling.., yet show different symmetry of tail associaton. Asymmetrically tail associated cases are perfectly correlated in the lower (A)  or upper (C) halfs of the distribution. The variables $(b_1,b_2)$ are transformed by multipling by $\sigma$ and then adding $(\mu_1,\mu_2)$.
\label{fig:noise}}
\end{figure}

\begin{figure}
\caption{%Pedagogical figure somehow displaying the different outcomes of competition in ATA and symmetric cases for an example. This will be cited from the Intro, so needs to be understandable to someone who does not know any of the ideas we introduce yet. So in particular you cannot use IGR as your metric of coexistence, you need to use something more intuitive. You will pick model parameter values and do two sims, one with ATA noise and one with symmtric noise, and show different outcomes. We should discuss and possibly take several iterations to get this figure right since it's a pedagogically important figure - if the reader is not convinced in the Intro that we are studying something that makes a difference, they are not going to slog through the Theory section, which is inevitably going to be tough going for many readers.
Populations inlfuenced by asymmetric tail association...
recover from low densities less often/ spend more time on average under one-species dominance. A) is a simulation over time of a population within a two species lottery model whose environmental response and competition is influenced by symmetric bivariate noise. B) is like A), but the population is influenced by bivariate noise with left-tailed asymmetric association. Both simulations were ran for total time of 100,000. Noticeable population density was defined at a threshold of 0.95 of total community space. Periods of noticable coexistence are marked in grey. While the population in B) had longer mean periods of coexistence, it had much longer mean periods of one-species dominace. The population in B) also recovered from low densities less often and thus has a lower ability to recover when rare and thus that community has lower coexistence.     
\label{fig:pedagog}}
\end{figure}

>>>>

To further explore the model and acheive G3 from the introduction i.e., identifying the ATA rescue or ATA exlusion subset of species trait space for which species will coexist or become competitively excluded in our model, we plotted the invasion growth rate [IGR], $r_{i/i}-q_{12}r_{i/j}$ and the IGR without ATA contributribution  $r_{i/i}-q_{12}r_{i/j} - \Delta_i^{[EC]}$ against the difference between species, $\mu_1-\mu_2$. The same mathematical simplifications described above hold, such that $\mu1$ and $\mu2$ are the only ways that the species differ and computations for coexistence in the lottery model only depend on the difference, $\mu_1-\mu_2$. We henceforward call this difference the trait difference between species or species trait difference or trait space subset range  or species trait range for short. Letting the trait $\mu$ vary within a one dimensional space, this space is then a trait space. Letting there be two points $\mu_1$ and $\mu_2$ within that trait space, then the difference, $\mu_1-\mu_2$, is a range of a subset of that trait space. We then define $\psi_{ATA}$ to be the subset of trait space that allows a postive IGR when ATA is included and $\psi_{~ATA}$ to be the subset of trait space that allows a postive IGR when ATA is not included. We then use symmetric difference $\psi_{ATA} \Delta \psi_{~ATA}$ to make the following conslusions. If the length/range?/rank? of $\psi_{ATA}$ > the length/range?/rank? of $\psi_{~ATA}$ then $\psi_{ATA} \Delta \psi_{~ATA}$ consitutes ATA rescue and the degree of species trait difference that allows ATA rescue to act, hence, the ATA rescue trait space. If the length/range/rank of $\psi_{ATA}$ < the length/range?/rank? of $\psi_{~ATA}$ then $\psi_{ATA} \Delta \psi_{~ATA}$ consitutes ATA exclusion and the degree of species trait difference that allows ATA exclusion to act, hence, the ATA exclusion trait space. In other words, ATA rescue is a subset of all $\mu_1-\mu_2$ that results in coexistence with ATA but failure to coexist without ATA and ATA exclusion is a subset of all $\mu_1-\mu_2$ that results in coexistence without ATA, but failure to coexist with ATA. We highlight and color code these areas in our plots. We made the plots for values of $\delta=0.5, 0.8, 1.0$ and $\sigma=2.0,4.0,5.0, 6.0$. The spaces in between the lines where they intersect the x-axis at y=0, indicate either an ATA rescue trait space or an ATA exclusion trait space. If the line plotting IGR is to the left of the line plotting IGR without ATA, then this space is an ATA rescue trait space. If the line plotting IGR is to the right of the line plotting IGR without ATA, then this space is an ATA exclusion trait space.

@@

\begin{figure}
\includegraphics[width=\textwidth]{"../results_figs/fig1.png"}
\caption{
%Figure displaying the noise. Three panels, left-, symmetric, and right-tail assoc noise. Use standard normal marginals.Explain in caption that alternatives for $(b_1,b_2)$ are obtained by multipling by $\sigma$ and then adding $(\mu_1,\mu_2)$.Jasmin to make and insert the figure here using the includegraphics command, and then draft a caption along the lines described here.
Simple example of bivariate noise with standard normal marginals. A) is left-tail associated, B) is symmetric, and C) is right-tail associated. All three cases have the same Pearson correlation of $\rho=$, up to sampling.., yet show different symmetry of tail associaton. Asymmetrically tail associated cases are perfectly correlated in the lower (A)  or upper (C) halfs of the distribution. The variables $(b_1,b_2)$ are transformed by multipling by $\sigma$ and then adding $(\mu_1,\mu_2)$.
\label{fig:noise}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{"../results_figs/figure2.PNG"}
\caption{%Pedagogical figure somehow displaying the different outcomes of competition in ATA and symmetric cases for an example. This will be cited from the Intro, so needs to be understandable to someone who does not know any of the ideas we introduce yet. So in particular you cannot use IGR as your metric of coexistence, you need to use something more intuitive. You will pick model parameter values and do two sims, one with ATA noise and one with symmtric noise, and show different outcomes. We should discuss and possibly take several iterations to get this figure right since it's a pedagogically important figure - if the reader is not convinced in the Intro that we are studying something that makes a difference, they are not going to slog through the Theory section, which is inevitably going to be tough going for many readers.
Populations inlfuenced by asymmetric tail association...
recover from low densities less often/ spend more time on average under one-species dominance. A) is a simulation over time of a population within a two species lottery model whose environmental response and competition is influenced by symmetric bivariate noise. B) is like A), but the population is influenced by bivariate noise with left-tailed asymmetric association. Both simulations were ran for total time of 100,000. Noticeable population density was defined at a threshold of 0.95 of total community space. Periods of noticable coexistence are marked in grey. While the population in B) had longer mean periods of coexistence, it had much longer mean periods of one-species dominace. The population in B) also recovered from low densities less often and thus has a lower ability to recover when rare and thus that community has lower coexistence.     
\label{fig:pedagog}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{"../results_figs/fig3_legend.pdf"}
\caption{The contribution of ATA is non-negligible. All panels show the breakdown of contributing mechanisms of coexistence in accordance to \cite{Ellner_2019} new decomposition method, against increasing $\sigma$. Each panel is defined by a unique set of parameters $\mu_1-\mu_2$ and $\delta$. 
\label{fig:fig3}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{"../results_figs/fig4_samexaxis50legend.pdf"}
\caption{ATA rescue and ATA exclusion can be defined within subsets species trait difference spaces. IGR (ATA included; yellow) and IGR without ATA (violet) are plotted against species trait difference $\mu_1-\mu_2$ in order to demonstrait ATA rescue and ATA exclusion. The red line, $\Delta_i^{[E \nat C]}$, is the difference between the two IGR's. Each panel is defined by a unique set of parameters $\sigma$ and $\delta$. ATA rescue is marked by regions in the panels where IGR wihtout ATA contribution is less than IGR at zero, denoted by regions of dotted yellow. ATA exclusion is marked by regions where IGR without ATA contribution is greater than IGR at zero, denoted by regions of dotted pink. ATA rescue is only seen in cases where $\delta=1$ (left column). All other cases shown here have subsets of species trait difference space where ATA exclusion occurs. Despite panels in the top row not containing the color coded region, these cases still contain ATA exlusion spaces; however, the space is too small to appropriately denote while still maintaining consistent x-axes across all panels. For $\delta<1$, ATA exclusion space increases down rows as $\sigma$ increases. ATA exclusion space also seems to increase across columns as $\delta$ decreases; however this trend is broken in the last row, indicating a more compex relationship between ATA exclusion space, $\delta$, and $\sigma$.  
\label{fig:fig4}}
\end{figure}
